---
title: "Tucson's Vanishing Winters"
subtitle: "Watch the Blues Fade and Reds Take Over"
author: "Vanishree Venkatachalapathy"
date: "03-18-2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    fig-width: 10
    fig-height: 5
    fig-align: center
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

Tucson, Arizona sits in the Sonoran Desert — a city where summer heat is a given, but where mild winters have long been part of the landscape's character. Over the past four decades, that balance has quietly shifted.

This project draws on 41 years of monthly NASA satellite temperature data (1984–2024) to map exactly how and when that shift occurred. Rather than a single global baseline, each month is compared against its own 41-year average — so a warm January is judged on January's terms, not July's. The result is a clear signal: the blue of cool, below-average months is fading, and the red of anomalous warmth is spreading across more months, more seasons, and more years.

> **Data:** NASA POWER Agroclimatology API · Monthly mean 2-metre air temperature (T2M) · Tucson, AZ (32.22°N, 110.97°W) · 1984–2024

---

## Methodology

The analysis follows a four-step pipeline: load packages, retrieve data from the NASA POWER API, reshape to long format, and compute per-month anomalies. All code is reproducible — expand any block below to inspect a step.

### Packages & Dependencies

```{r}
#| label: setup

library(nasapower)   # NASA POWER API client — get_power()
library(dplyr)       # Data wrangling
library(tidyr)       # pivot_longer, pivot_wider
library(lubridate)   # Date handling — make_date()
library(jsonlite)    # toJSON() — bridge R data to JavaScript
library(htmltools)   # tagList, tags$*, browsable() — embed HTML/JS in Quarto
library(ggplot2)     # Static visualizations
library(ggtext)      # element_markdown() for rich text in ggplot2 titles
library(glue)        # String interpolation for annotations
library(knitr)       # kable() for summary tables
library(scales)      # Label formatting in ggplot2
```

---

### Data Source

Monthly mean 2-metre air temperature for Tucson, Arizona was retrieved from the NASA POWER agroclimatology API, covering the full 41-year period 1984–2024. The chunk is cached to avoid redundant API calls on re-render.

```{r}
#| label: data-acquisition
#| cache: true

raw <- get_power(
  community    = "ag",                           # agroclimatology dataset
  lonlat       = c(-110.97, 32.22),             # Tucson, AZ (longitude, latitude)
  pars         = "T2M",                          # 2-metre mean air temperature (°C)
  dates        = c("1984-01-01", "2024-12-31"), # full 41-year record
  temporal_api = "monthly"                       # monthly resolution
)

# Preview the raw API response structure
head(raw, 14) |> kable(caption = "Raw NASA POWER response (first 14 rows)")
```

---

### Data Preparation

We reshape the wide API response to long format, one row per (year, month) observation, and construct a proper `date` column for time-series plotting.

```{r}
#| label: data-cleaning

monthly <- raw |>
  dplyr::select(year = YEAR, JAN, FEB, MAR, APR, MAY, JUN,
                JUL, AUG, SEP, OCT, NOV, DEC) |>
  tidyr::pivot_longer(
    cols      = JAN:DEC,
    names_to  = "month_abb",
    values_to = "temp"
  ) |>
  mutate(
    month = match(tolower(month_abb), tolower(month.abb)),
    date  = lubridate::make_date(year, month, 1L)
  ) |>
  dplyr::select(date, year, month, temp) |>
  arrange(date)

# Confirm: should be exactly 492 rows (41 years × 12 months)
cat("Total observations:", nrow(monthly), "\n")
cat("Date range:", format(min(monthly$date)), "to", format(max(monthly$date)), "\n")
```

---

### Temperature Anomaly Calculation

Each month's anomaly is its observed temperature minus that **month's own** 41-year mean. This preserves the seasonal structure of the signal: a warm January is judged against January's average, so no month is unfairly penalized or rewarded by comparison to a different season.

```{r}
#| label: anomaly-computation

# Per-month baseline: each of the 12 months gets its own 41-year mean
monthly_baseline <- monthly |>
  group_by(month) |>
  summarise(baseline_temp = mean(temp, na.rm = TRUE), .groups = "drop")

# Anomaly = observed temp minus that month's baseline
monthly_anomaly <- monthly |>
  left_join(monthly_baseline, by = "month") |>
  mutate(anomaly = temp - baseline_temp) |>
  dplyr::select(year, month, temp, baseline_temp, anomaly)

# Export anomaly data to JSON — injected into the JavaScript particle field
# as ANOMALY_DATA: an array of {year, month, anomaly} objects (492 rows total)
anomaly_json <- jsonlite::toJSON(
  monthly_anomaly |> dplyr::select(year, month, anomaly),
  dataframe = "rows",
  digits    = 4
)

# Summary table: per-month baselines
monthly_baseline |>
  mutate(month_name = month.abb[month]) |>
  dplyr::select(Month = month_name, `Baseline Mean Temp (°C)` = baseline_temp) |>
  kable(
    caption = "41-year per-month baseline temperatures, Tucson AZ (1984–2024)",
    digits  = 2
  )
```

The overall warming signal becomes clear when we summarise at the annual level:

```{r}
#| label: annual-summary-table

# Collapse 492 monthly anomalies to one mean anomaly per year (41 rows)
annual_summary <- monthly_anomaly |>
  group_by(year) |>
  summarise(
    mean_anomaly = mean(anomaly, na.rm = TRUE),
    .groups = "drop"
  )

cat("Total warming (mean anomaly first vs last decade):\n")
early <- mean(annual_summary$mean_anomaly[annual_summary$year <= 1993]) # first decade mean
late  <- mean(annual_summary$mean_anomaly[annual_summary$year >= 2015]) # last decade mean
cat(glue("  1984–1993 mean anomaly: {round(early, 2)}°C\n"))
cat(glue("  2015–2024 mean anomaly: {round(late,  2)}°C\n"))
cat(glue("  Shift: +{round(late - early, 2)}°C\n"))

hottest  <- annual_summary |> slice_max(mean_anomaly, n = 1)
coolest  <- annual_summary |> slice_min(mean_anomaly, n = 1)
cat(glue("\nHottest year:  {hottest$year}  (+{round(hottest$mean_anomaly,  2)}°C)\n"))
cat(glue("Coolest year:  {coolest$year}  ({round(coolest$mean_anomaly, 2)}°C)\n"))

# Export hottest/coolest for JS annotation in particle field
hottest_json <- jsonlite::toJSON(list(year=hottest$year, anom=round(hottest$mean_anomaly,2)), auto_unbox=TRUE)
coolest_json <- jsonlite::toJSON(list(year=coolest$year, anom=round(coolest$mean_anomaly,2)), auto_unbox=TRUE)
```

---

## The Warming Signal

Four complementary views of the same 41-year record. Together they build the case before the interactive visualization: Tucson is warming, it is accelerating, and it is happening across every season.

### Warming Stripes

One stripe per year, colored by its annual mean temperature anomaly. No axes, no numbers — just the pattern. Read it left to right.

```{r}
#| label: warming-stripes
#| fig-height: 3

annual_summary |>
  ggplot(aes(x = year, y = 1, fill = mean_anomaly)) +
  geom_col(width = 1, show.legend = FALSE) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "#f7f7f7",
    high     = "#b2182b",
    midpoint = 0
  ) +
  scale_x_continuous(
    breaks = c(1984, 1990, 2000, 2010, 2020, 2024),
    expand = c(0, 0)
  ) +
  theme_void() +
  theme(
    axis.text.x  = element_text(color = "grey60", size = 8, margin = margin(t = 4)),
    axis.ticks.x = element_line(color = "grey60"),
    plot.title   = element_markdown(size = 13, face = "bold", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 9, color = "grey50", margin = margin(b = 8)),
    plot.margin  = margin(10, 10, 10, 10)
  ) +
  labs(
    title    = "**Tucson Warming Stripes** · Annual mean temperature anomaly",
    subtitle = "Each stripe = one year · Blue = cooler than baseline · Red = warmer than baseline · Baseline: 1984–2024 mean"
  )
```

The pattern is stark. Before 2000, blue dominated the timeline. After 2012, warm anomalies have been nearly unbroken — only a handful of years since then register below the long-term baseline.

### Annual Anomaly Trend

Annual mean anomalies shown as bars, with a LOESS smoother overlaid to reveal the underlying trend. The hottest and coolest years are annotated directly on the chart.

```{r}
#| label: anomaly-trend
#| fig-height: 4.5

hottest_label <- annual_summary |> slice_max(mean_anomaly, n = 1)
coolest_label <- annual_summary |> slice_min(mean_anomaly, n = 1)

annual_summary |>
  ggplot(aes(x = year, y = mean_anomaly)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.4, linetype = "dashed") +
  geom_col(aes(fill = mean_anomaly), width = 0.7, show.legend = FALSE) +
  geom_smooth(method = "loess", span = 0.5, se = FALSE,
              color = "white", linewidth = 0.9, linetype = "solid") +
  geom_label(
    data  = hottest_label,
    aes(label = glue("{year}\n+{round(mean_anomaly,2)}°C")),
    size  = 2.8, fill = "#b2182b", color = "white",
    label.padding = unit(0.3, "lines"), label.size = 0
  ) +
  geom_label(
    data  = coolest_label,
    aes(label = glue("{year}\n{round(mean_anomaly,2)}°C")),
    size  = 2.8, fill = "#2166ac", color = "white",
    label.padding = unit(0.3, "lines"), label.size = 0
  ) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0) +
  scale_x_continuous(breaks = seq(1984, 2024, by = 4)) +
  scale_y_continuous(labels = function(x) paste0(ifelse(x > 0, "+", ""), x, "°C")) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor  = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title        = element_markdown(size = 13, face = "bold"),
    plot.subtitle     = element_text(size = 9, color = "grey50"),
    axis.title        = element_text(size = 9, color = "grey50")
  ) +
  labs(
    title    = "**Annual Temperature Anomaly** · Tucson, AZ · 1984–2024",
    subtitle = "Bars = annual mean anomaly vs 41-yr baseline · White line = LOESS smoother (span = 0.5)",
    x        = NULL,
    y        = "Anomaly (°C)"
  )
```

The LOESS smoother reveals a persistent upward trajectory. The warmest years on record cluster almost entirely in the most recent decade — a signal of accelerating, not merely steady, change.

### Monthly Heat Calendar

Each cell is one (month, year) combination. Reading down a column shows how an entire year's months compare to baseline; reading across a row shows how a particular month has trended over time.

```{r}
#| label: monthly-heat-calendar
#| fig-height: 5

monthly_anomaly |>
  mutate(month_fct = factor(month.abb[month], levels = rev(month.abb))) |>
  ggplot(aes(x = year, y = month_fct, fill = anomaly)) +
  geom_tile(color = "grey15", linewidth = 0.15) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "#f7f7f7",
    high     = "#b2182b",
    midpoint = 0,
    name     = "Anomaly (°C)",
    limits   = c(-3, 3),
    oob      = scales::squish
  ) +
  scale_x_continuous(
    breaks = c(1984, 1990, 2000, 2010, 2020, 2024),
    expand = c(0, 0)
  ) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid    = element_blank(),
    legend.position = "right",
    legend.key.height = unit(1.2, "cm"),
    plot.title    = element_markdown(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 9, color = "grey50"),
    axis.text.y   = element_text(size = 8)
  ) +
  labs(
    title    = "**Monthly Temperature Anomaly Calendar** · Tucson, AZ · 1984–2024",
    subtitle = "Each tile = one month · Color = departure from that month's 41-yr baseline",
    x        = NULL,
    y        = NULL
  )
```

Read down any recent year's column and red has spread well beyond the summer months into spring and autumn. The cool blue winters that once anchored the calendar are becoming increasingly rare.

### Monthly Warming Rose

Each petal represents one calendar month, with its length showing the average anomaly for that month across all 41 years. This view reveals the seasonal distribution of warming — which months have shifted most relative to their own baselines.

```{r}
#| label: rose-chart
#| fig-height: 5.5

monthly_anomaly |>
  group_by(month) |>
  summarise(mean_anom = mean(anomaly, na.rm = TRUE), .groups = "drop") |>
  mutate(
    month_name = factor(month.abb[month], levels = month.abb)
  ) |>
  ggplot(aes(x = month_name, y = mean_anom, fill = mean_anom)) +
  geom_col(width = 0.85, show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.3) +
  coord_polar(start = -pi/12) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0) +
  scale_y_continuous(labels = function(x) paste0(ifelse(x > 0, "+", ""), round(x, 1), "°C")) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.x = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 9, face = "bold"),
    axis.text.y        = element_text(size = 7, color = "grey60"),
    axis.title         = element_blank(),
    plot.title         = element_markdown(size = 13, face = "bold", hjust = 0.5),
    plot.subtitle      = element_text(size = 9, color = "grey50", hjust = 0.5)
  ) +
  labs(
    title    = "**Monthly Warming Rose** · Tucson, AZ · 1984–2024",
    subtitle = "Petal length = mean anomaly for that calendar month across all 41 years"
  )
```

No month has been spared. While summer months show the largest absolute anomalies, even December and January — Tucson's mildest months — are running consistently warmer than their historical averages.

---

## Explore 41 Years of Tucson Temperatures

The visualization below encodes all 492 month–year combinations simultaneously as a living particle field. Each column of particles is a year (1984–2024, left to right). Each row is a calendar month (January at top, December at bottom). A particle's **color** shows whether that month ran warmer or cooler than its 41-year average:

- **Red** — warmer than baseline
- **Blue** — cooler than baseline
- **Amber** — a warm anomaly in a summer month (June–August)

**Use the slider** at the bottom to scrub through years and highlight a column. **Hover over a month row** to inspect it and read the exact anomaly value in the stat box. Press **▶** to animate through all 41 years.

```{r}
#| label: particle-field
#| echo: true

library(htmltools)

# ── Output path ────────────────────────────────────────────────────────────
out_dir  <- "/Users/Vani/Documents/My R/Viz Challenge 2026/Data-Viz-Challenge-2026/Data Viz Challenge_Mar2026/Tucon temperatures/output"
out_file <- file.path(out_dir, "tucson_month_year_particle_field.html")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ── Build the HTML string (used for both embed + standalone file) ──────────
# anomaly_json was computed in Section 4 via jsonlite::toJSON()
particle_html_str <- paste0(
'
<!DOCTYPE html>
<html>
<head>
<meta charset="utf-8">
<title>Tucson \u00b7 Month \u00d7 Year Particle Field \u00b7 1984\u20132024</title>
<style>
* { margin:0; padding:0; box-sizing:border-box; }
body { background:#07070e; font-family:\'Georgia\',serif; overflow:hidden; }
canvas { display:block; position:absolute; top:0; left:0; }
#wrap { position:relative; width:100%; }
#border-frame {
  position:fixed; top:0; left:0; right:0; bottom:0;
  border:1px solid rgba(220,200,155,0.35);
  pointer-events:none; z-index:999;
}
#header {
  position:absolute; top:0; left:0; right:0; z-index:20;
  background:rgba(7,7,14,0.93);
  border-bottom:1px solid rgba(220,200,155,0.07);
  padding:9px 20px;
  display:flex; align-items:center; justify-content:space-between; gap:16px;
}
#header .title-block { display:flex; flex-direction:column; gap:1px; white-space:nowrap; }
#header .title-main { font-size:22px; font-weight:700; color:rgba(220,200,160,0.92); letter-spacing:0.01em; line-height:1.1; }
#header .title-sub  { font-size:10px; font-style:italic; color:rgba(220,200,160,0.82); letter-spacing:0.04em; margin-top:1px; }
#header .desc  { font-size:9.5px; color:rgba(220,200,160,0.82); line-height:1.5; text-align:center; flex:1; }
#header .keys  { display:flex; gap:14px; white-space:nowrap; }
#header .keys span {
  font-size:8.5px; letter-spacing:0.1em; text-transform:uppercase;
  display:flex; align-items:center; gap:6px; color:rgba(220,200,160,0.82);
}
#header .keys span::before { content:\'\'; width:18px; height:2.5px; border-radius:1px; flex-shrink:0; }
.k-warm::before { background:#e05530; }
.k-cool::before { background:#5588e0; }
.k-sum::before  { background:#e8a020; }
#ui {
  position:absolute; bottom:0; left:0; right:0; z-index:20;
  padding:8px 22px 12px;
  background:linear-gradient(transparent, rgba(7,7,14,0.97) 36%);
  display:flex; align-items:center; gap:18px;
}
.stat { text-align:center; min-width:90px; }
.stat .val { font-size:14px; font-weight:700; }
.stat .lbl { font-size:7px; letter-spacing:0.12em; text-transform:uppercase; color:rgba(220,200,160,0.82); margin-top:2px; line-height:1.4; }
#slider-wrap { flex:1; display:flex; flex-direction:column; gap:4px; }
#slider-wrap label { font-size:7.5px; letter-spacing:0.13em; text-transform:uppercase; color:rgba(220,200,160,0.82); text-align:center; }
input[type=range] {
  width:100%; height:4px; cursor:pointer;
  -webkit-appearance:none; appearance:none; border-radius:2px; outline:none; border:none;
  background:linear-gradient(to right,#1a3388,#882211);
}
input[type=range]::-webkit-slider-thumb {
  -webkit-appearance:none; width:14px; height:14px; border-radius:50%;
  background:#c8b080; border:2px solid rgba(220,195,150,0.5); cursor:pointer;
}
#play-btn {
  width:38px; height:38px; border-radius:50%; flex-shrink:0;
  background:rgba(7,7,14,0.6);
  border:1.5px solid rgba(220,200,160,0.5);
  color:rgba(220,200,160,0.88); font-size:14px;
  cursor:pointer; display:flex; align-items:center; justify-content:center;
  transition:border-color 0.2s, color 0.2s;
}
#play-btn:hover { border-color:rgba(220,200,160,0.9); color:rgba(255,240,200,1); }
#bar-panel {
  position:absolute; left:0; right:0; z-index:15;
  background:rgba(7,7,14,0.88);
  border-top:1px solid rgba(220,200,155,0.07);
  border-bottom:1px solid rgba(220,200,155,0.06);
}
#bar-c { display:block; width:100%; }
#month-gutter {
  position:absolute; left:0; z-index:18;
  width:48px;
  background:linear-gradient(to right, rgba(7,7,14,0.97) 70%, rgba(7,7,14,0));
  display:flex; flex-direction:column; pointer-events:none;
}
#month-gutter .m-label {
  display:flex; align-items:center;
  padding-left:8px;
  font-family:\'Georgia\',serif;
  font-size:11px;
  color:rgba(220,200,160,0.88);
  letter-spacing:0.04em;
  flex-shrink:0;
  transition: font-size 0.15s ease, color 0.15s ease;
}
</style>
</head>
<body>
<div id="border-frame"></div>
<div id="wrap">
  <canvas id="c"></canvas>
  <div id="month-gutter">
    <div class="m-label">Jan</div>
    <div class="m-label">Feb</div>
    <div class="m-label">Mar</div>
    <div class="m-label">Apr</div>
    <div class="m-label">May</div>
    <div class="m-label">Jun</div>
    <div class="m-label">Jul</div>
    <div class="m-label">Aug</div>
    <div class="m-label">Sep</div>
    <div class="m-label">Oct</div>
    <div class="m-label">Nov</div>
    <div class="m-label">Dec</div>
  </div>
  <div id="bar-panel">
    <canvas id="bar-c"></canvas>
  </div>
  <div id="header">
    <div class="title-block">
      <span class="title-main">Tucson\u2019s Vanishing Winters</span>
      <span class="title-sub">Watch the Blues Fade and Reds Take Over</span>
    </div>
    <div class="desc">
      Y-axis = month of year (Jan at top \u2192 Dec at bottom) \u00b7 X-axis = year (1984 \u2192 2024)<br>
      Each particle sits in its month row and year column \u00b7 color = temperature anomaly vs 41-yr baseline
    </div>
    <div class="keys">
      <span class="k-warm">Warmer than baseline</span>
      <span class="k-cool">Cooler than baseline</span>
      <span class="k-sum">Summer peak (Jun\u2013Aug)</span>
    </div>
  </div>
  <div id="ui">
    <button id="play-btn" title="Play / Pause">\u25b6</button>
    <div id="slider-wrap">
      <label>\u25c4 drag to highlight a year \u25ba</label>
      <input type="range" id="slider" min="1984" max="2024" value="1984" step="1">
      <label id="yr-lbl" style="color:rgba(220,200,160,0.88); font-size:9px; letter-spacing:0.1em;">
        Year: 1984 \u2014 hover a month row to inspect
      </label>
    </div>
    <div class="stat">
      <div class="val" id="stat-anom" style="color:#e05530">\u2014</div>
      <div class="lbl">Anomaly for<br>highlighted cell</div>
    </div>
  </div>
</div>
<script>
// ── Data bridge: R exports these via jsonlite::toJSON(), injected here as JS constants ──
const ANOMALY_DATA = ',
anomaly_json,
'; // array of {year, month, anomaly} — 492 rows (41 years × 12 months)
const HOTTEST = ',
hottest_json,
'; // {year, anom} — used to annotate the hottest year column
const COOLEST = ',
coolest_json,
'; // {year, anom} — used to annotate the coolest year column

// Look up anomaly for a given month (0-indexed) and year
function monthAnom(month0, year) {
  const row = ANOMALY_DATA.find(d => d.month === month0 + 1 && d.year === year);
  return row ? row.anomaly : 0;
}
const canvas   = document.getElementById("c");
const ctx      = canvas.getContext("2d");
const barCanvas= document.getElementById("bar-c");
const barCtx   = barCanvas.getContext("2d");
const wrap     = document.getElementById("wrap");
const barPanel = document.getElementById("bar-panel");
const gutter   = document.getElementById("month-gutter");
const gutterLabels = gutter.querySelectorAll(".m-label");
const slider   = document.getElementById("slider");
const HEADER_H = 72, UI_H = 66, BAR_H = 160; // fixed layout heights in CSS pixels
const MONTHS = ["Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"];
const YEARS  = Array.from({length:41}, (_,i) => 1984+i);
let W, FIELD_H, ROW_H, COL_W;
function resize() {
  W       = window.innerWidth;
  FIELD_H = window.innerHeight - HEADER_H - UI_H - BAR_H;
  ROW_H   = FIELD_H / 12;
  COL_W   = W / 41;
  wrap.style.height      = (FIELD_H + HEADER_H + BAR_H + UI_H) + "px";
  canvas.style.top       = HEADER_H + "px";
  canvas.width           = W * devicePixelRatio;
  canvas.height          = FIELD_H * devicePixelRatio;
  canvas.style.width     = W + "px";
  canvas.style.height    = FIELD_H + "px";
  barPanel.style.top     = (HEADER_H + FIELD_H) + "px";
  barPanel.style.height  = BAR_H + "px";
  barCanvas.width        = W * devicePixelRatio;
  barCanvas.height       = BAR_H * devicePixelRatio;
  barCanvas.style.height = BAR_H + "px";
  // gutter: align with particle canvas
  gutter.style.top    = HEADER_H + "px";
  gutter.style.height = FIELD_H + "px";
  gutterLabels.forEach(el => { el.style.height = (FIELD_H/12) + "px"; });
}
resize();
window.addEventListener("resize", () => { resize(); boot(); drawBarChart(); });
// ── Fractional Brownian Motion noise (hand-coded, no external library) ──────
// hash(): pseudo-random number from a single integer using sine trick
function hash(n) { return Math.sin(n*127.1)*43758.5453 - Math.floor(Math.sin(n*127.1)*43758.5453); }
// smoothN(): bilinear interpolation of hash values on a 2D grid (value noise)
function smoothN(x, y) {
  const ix=Math.floor(x), iy=Math.floor(y), fx=x-ix, fy=y-iy;
  const ux=fx*fx*(3-2*fx), uy=fy*fy*(3-2*fy); // smoothstep curve for softer blending
  const a=hash(ix+iy*57), b=hash(ix+1+iy*57), c=hash(ix+(iy+1)*57), d=hash(ix+1+(iy+1)*57);
  return a+(b-a)*ux+(c-a)*uy+(d-b-c+a)*ux*uy;
}
// fbm(): layered noise at 3 frequencies — each layer adds finer detail (octaves)
function fbm(x, y, seed=0) {
  return smoothN(x+seed,y+seed)*0.5 + smoothN(x*2+seed,y*2+seed)*0.25 + smoothN(x*4,y*4)*0.125;
}
const NP = 1400; // total number of particles in the field
let particles = [];
// Create one particle assigned to a random (month, year) cell
function makeParticle() {
  const month0   = Math.floor(Math.random()*12);
  const year     = YEARS[Math.floor(Math.random()*YEARS.length)];
  const anom     = monthAnom(month0, year);
  const xc       = (YEARS.indexOf(year)+0.5)*COL_W;
  const yc       = (month0+0.5)*ROW_H;
  const isSummer = month0 >= 5 && month0 <= 7;
  return {
    month0, year, anom, xc, yc, isSummer,
    x: xc+(Math.random()-0.5)*COL_W*1.1, y: yc+(Math.random()-0.5)*ROW_H*0.85,
    vx:0, vy:0, seed: month0*3.7, age: Math.random()*110,
    maxAge: 80+Math.random()*140,
    size: (isSummer?0.9:0.65)+Math.random()*1.3*Math.min(1,Math.abs(anom)*0.9+0.3), // larger dot = stronger anomaly
    speed: 0.35+Math.random()*0.75,
  };
}
function boot() { particles = Array.from({length:NP}, makeParticle); }
// Maps anomaly value to an RGBA color string:
//   warm summer (Jun–Aug) → amber, warm non-summer → red, cool → blue
function anomColor(anom, isSummer, alpha) {
  const v=Math.max(-1.5,Math.min(1.5,anom)); let r,g,b;
  if(isSummer&&v>0){ r=Math.round(220+Math.min(35,v*25)); g=Math.round(140-v*55); b=10; } // amber
  else if(v>0){ r=Math.round(180+v*55); g=Math.round(50-v*30); b=15; }                    // red
  else{ const f=-v; r=15; g=Math.round(75+f*55); b=Math.round(185+f*65); }                // blue
  return "rgba("+r+","+g+","+b+","+alpha+")";
}
let highlightYear=1984, hoverMonth=null;
let isPlaying=false, playTimer=null;
const playBtn=document.getElementById("play-btn");
function startPlay(){
  isPlaying=true; playBtn.textContent="\u23f8";
  playTimer=setInterval(()=>{
    let yr=parseInt(slider.value)+1;
    if(yr>2024) yr=1984;
    slider.value=yr; updateYear();
  },400);
}
function stopPlay(){
  isPlaying=false; playBtn.textContent="\u25b6";
  clearInterval(playTimer);
}
playBtn.addEventListener("click",()=>{ isPlaying?stopPlay():startPlay(); });
function fmt(v){ return (v>=0?"+":"")+v.toFixed(2)+"\u00b0C"; }
function updateYear() {
  highlightYear=parseInt(slider.value);
  document.getElementById("yr-lbl").textContent="Year: "+highlightYear+(hoverMonth!==null?" \u00b7 "+MONTHS[hoverMonth]:" \u2014 hover a month row to inspect");
  updateStatBox();
  drawBarChart();
}
function updateStatBox() {
  if(hoverMonth!==null){
    const a=monthAnom(hoverMonth,highlightYear);
    document.getElementById("stat-anom").textContent=fmt(a);
    document.getElementById("stat-anom").style.color=a<0?"#5588e0":(hoverMonth>=5&&hoverMonth<=7)?"#e8a020":"#e05530";
  } else {
    document.getElementById("stat-anom").textContent="\u2014";
  }
}
slider.addEventListener("input", updateYear);
canvas.addEventListener("mousemove", e => {
  const rect=canvas.getBoundingClientRect(), y=e.clientY-rect.top;
  hoverMonth=Math.floor(y/ROW_H);
  if(hoverMonth<0||hoverMonth>11) hoverMonth=null;
  updateStatBox();
  document.getElementById("yr-lbl").textContent="Year: "+highlightYear+(hoverMonth!==null?" \u00b7 "+MONTHS[hoverMonth]:" \u2014 hover a month row to inspect");
  drawBarChart();
});
canvas.addEventListener("mouseleave", () => { hoverMonth=null; updateStatBox(); drawBarChart(); });
',
'function drawBarChart() {
  const dpr=devicePixelRatio;
  const cW=barCanvas.width, cH=barCanvas.height;
  barCtx.clearRect(0,0,cW,cH);
  const PAD_L=44*dpr, PAD_R=16*dpr, PAD_T=28*dpr, PAD_B=26*dpr;
  const MONTH_LABEL_H=36*dpr;
  const chartH=cH-PAD_T-PAD_B-MONTH_LABEL_H;
  const chartW=cW-PAD_L-PAD_R;
  const anoms=MONTHS.map((m,i)=>({label:m,anom:monthAnom(i,highlightYear),month0:i}));
  const maxAbs=Math.max(1.5,...anoms.map(d=>Math.abs(d.anom)));
  const barW=chartW/12;
  const zeroY=PAD_T+chartH*0.5;
  // monthLabelY: always below deepest possible anomaly label (zeroY + chartH/2 + 11 + 9 px font + 6 gap)
  const monthLabelY=PAD_T+chartH+26*dpr;
  // section title left only
  barCtx.fillStyle="rgba(220,200,160,0.88)";
  barCtx.font="bold "+(8*dpr)+"px Georgia"; barCtx.textAlign="left";
  barCtx.fillText("MONTHLY ANOMALY PROFILE",PAD_L,16*dpr);
  // hottest/coldest year labels — sit at top of bar panel, below the particle field axis
  const hottestBX=((YEARS.indexOf(HOTTEST.year)+0.5)*COL_W)*devicePixelRatio;
  barCtx.fillStyle="rgba(220,55,25,0.92)"; barCtx.font=(7.5*dpr)+"px Georgia"; barCtx.textAlign="center";
  barCtx.fillText("hottest year", hottestBX, 10*dpr);
  const coolestBX=((YEARS.indexOf(COOLEST.year)+0.5)*COL_W)*devicePixelRatio;
  barCtx.fillStyle="rgba(40,100,220,0.92)";
  barCtx.fillText("coldest year", coolestBX, 10*dpr);
  // subtle grid lines + y-axis labels
  [-maxAbs,0,maxAbs].forEach(t=>{
    const ty=PAD_T+chartH*(1-(t+maxAbs)/(2*maxAbs));
    barCtx.strokeStyle="rgba(255,255,255,0.05)"; barCtx.lineWidth=0.5*dpr;
    barCtx.setLineDash([3*dpr,5*dpr]);
    barCtx.beginPath(); barCtx.moveTo(PAD_L,ty); barCtx.lineTo(PAD_L+chartW,ty); barCtx.stroke();
    barCtx.setLineDash([]);
    barCtx.fillStyle="rgba(220,200,160,0.82)"; barCtx.font=(7*dpr)+"px Georgia"; barCtx.textAlign="right";
    const lbl=t===0?"0":(t>0?"+":"")+t.toFixed(1)+"\u00b0C";
    barCtx.fillText(lbl,PAD_L-5*dpr,ty+3*dpr);
  });
  // zero line
  barCtx.strokeStyle="rgba(200,175,130,0.25)"; barCtx.lineWidth=0.8*dpr;
  barCtx.beginPath(); barCtx.moveTo(PAD_L,zeroY); barCtx.lineTo(PAD_L+chartW,zeroY); barCtx.stroke();
',
'  // bars
  anoms.forEach((d,i)=>{
    const isSummer=d.month0>=5&&d.month0<=7;
    const v=Math.max(-1.5,Math.min(1.5,d.anom)); let r,g,b;
    if(isSummer&&v>0){r=Math.round(220+Math.min(35,v*25));g=Math.round(140-v*55);b=10;}
    else if(v>0){r=Math.round(180+v*55);g=Math.round(50-v*30);b=15;}
    else{const f=-v;r=15;g=Math.round(75+f*55);b=Math.round(185+f*65);}
    const isHov=(hoverMonth===d.month0);
    const alpha=isHov?1.0:0.72;
    const col="rgba("+r+","+g+","+b+","+alpha+")";
    const colFull="rgba("+r+","+g+","+b+",1)";
    const barPx=(Math.abs(d.anom)/(2*maxAbs))*chartH;
    const x=PAD_L+i*barW;
    const y=d.anom>=0?zeroY-barPx:zeroY;
    const cx=x+barW/2;
    // bar fill
    barCtx.fillStyle=col;
    barCtx.fillRect(x+barW*0.1,y,barW*0.8,barPx);
    // highlight border
    if(isHov){
      barCtx.strokeStyle="rgba(220,195,130,0.75)";
      barCtx.lineWidth=1.2*dpr;
      barCtx.strokeRect(x+barW*0.1,y,barW*0.8,barPx);
    }
    // anomaly value — outer tip, bar color; larger + brighter on hover
    const anomSize=isHov?11*dpr:9*dpr;
    barCtx.font=(anomSize)+"px Georgia"; barCtx.textAlign="center";
    barCtx.fillStyle=colFull;
    const anomLbl=(d.anom>=0?"+":"")+d.anom.toFixed(2);
    const anomY=d.anom>=0 ? y-6*dpr : y+barPx+anomSize+4*dpr;
    barCtx.fillText(anomLbl,cx,anomY);
    // month label — pinned to bottom; larger + brighter on hover
    const mSize=isHov?13*dpr:10*dpr;
    barCtx.font="bold "+mSize+"px Georgia"; barCtx.textAlign="center";
    barCtx.fillStyle=isHov?"rgba(255,240,200,1)":"rgba("+r+","+g+","+b+",0.82)";
    barCtx.fillText(d.label,cx,monthLabelY);
  });
}
function drawGrid() {
  const dpr=devicePixelRatio; ctx.save();
  MONTHS.forEach((m,i) => {
    if(i%2===0){ ctx.fillStyle="rgba(255,255,255,0.012)"; ctx.fillRect(0,i*ROW_H*dpr,W*dpr,ROW_H*dpr); }
    if(i>=5&&i<=7){ ctx.fillStyle="rgba(200,130,10,0.025)"; ctx.fillRect(0,i*ROW_H*dpr,W*dpr,ROW_H*dpr); }
    if(i===hoverMonth){ ctx.fillStyle="rgba(220,195,130,0.04)"; ctx.fillRect(0,i*ROW_H*dpr,W*dpr,ROW_H*dpr); }
    ctx.strokeStyle="rgba(255,255,255,0.04)"; ctx.lineWidth=0.5*dpr;
    ctx.beginPath(); ctx.moveTo(0,i*ROW_H*dpr); ctx.lineTo(W*dpr,i*ROW_H*dpr); ctx.stroke();
    // update gutter label highlight state
    gutterLabels[i].style.color      = i===hoverMonth ? "rgba(255,240,200,1.0)" : "rgba(220,200,160,0.88)";
    gutterLabels[i].style.fontWeight = i===hoverMonth ? "700" : "400";
    gutterLabels[i].style.fontSize   = i===hoverMonth ? "17px" : "11px";
  });
  [1984,1990,2000,2010,2020,2024].forEach(yr => {
    const x=((YEARS.indexOf(yr)+0.5)*COL_W)*dpr;
    ctx.strokeStyle="rgba(200,175,120,0.08)"; ctx.lineWidth=0.5*dpr;
    ctx.setLineDash([2*dpr,6*dpr]);
    ctx.beginPath(); ctx.moveTo(x,0); ctx.lineTo(x,FIELD_H*dpr); ctx.stroke();
    ctx.setLineDash([]);
    ctx.fillStyle="rgba(160,135,85,0.3)"; ctx.font=(7.5*dpr)+"px Georgia"; ctx.textAlign="center";
    ctx.fillText(yr,x,(FIELD_H-3)*dpr);
  });
  const hx=((YEARS.indexOf(highlightYear)+0.5)*COL_W)*dpr;
  const hG=ctx.createLinearGradient(hx-15*dpr,0,hx+15*dpr,0);
  hG.addColorStop(0,"rgba(220,195,130,0)"); hG.addColorStop(0.5,"rgba(220,195,130,0.07)"); hG.addColorStop(1,"rgba(220,195,130,0)");
  ctx.fillStyle=hG; ctx.fillRect(hx-15*dpr,0,30*dpr,FIELD_H*dpr);
  ctx.strokeStyle="rgba(220,195,130,0.25)"; ctx.lineWidth=1*dpr;
  ctx.beginPath(); ctx.moveTo(hx,0); ctx.lineTo(hx,FIELD_H*dpr); ctx.stroke();
  // large year label for highlighted year — travels with slider
  ctx.fillStyle="rgba(220,195,130,0.92)"; ctx.font="bold "+(18*dpr)+"px Georgia"; ctx.textAlign="center";
  ctx.fillText(highlightYear,hx,(FIELD_H-14)*dpr);
  const triS=7*dpr;
  // ── Hottest year: red column glow + triangle + year ──
  const hottestX=((YEARS.indexOf(HOTTEST.year)+0.5)*COL_W)*dpr;
  const hotG=ctx.createLinearGradient(hottestX-COL_W*dpr,0,hottestX+COL_W*dpr,0);
  hotG.addColorStop(0,"rgba(200,40,15,0)"); hotG.addColorStop(0.5,"rgba(200,40,15,0.12)"); hotG.addColorStop(1,"rgba(200,40,15,0)");
  ctx.fillStyle=hotG; ctx.fillRect(hottestX-COL_W*dpr,0,COL_W*2*dpr,FIELD_H*dpr);
  ctx.fillStyle="rgba(220,55,25,0.95)";
  ctx.beginPath(); ctx.moveTo(hottestX,(FIELD_H-2)*dpr); ctx.lineTo(hottestX-triS,(FIELD_H-2+triS*1.6)*dpr); ctx.lineTo(hottestX+triS,(FIELD_H-2+triS*1.6)*dpr); ctx.closePath(); ctx.fill();
  ctx.font="bold "+(7.5*dpr)+"px Georgia"; ctx.textAlign="center";
  ctx.fillText(HOTTEST.year, hottestX, (FIELD_H-5)*dpr);
  // ── Coolest year: blue column glow + triangle + year ──
  const coolestX=((YEARS.indexOf(COOLEST.year)+0.5)*COL_W)*dpr;
  const coolG=ctx.createLinearGradient(coolestX-COL_W*dpr,0,coolestX+COL_W*dpr,0);
  coolG.addColorStop(0,"rgba(30,90,210,0)"); coolG.addColorStop(0.5,"rgba(30,90,210,0.12)"); coolG.addColorStop(1,"rgba(30,90,210,0)");
  ctx.fillStyle=coolG; ctx.fillRect(coolestX-COL_W*dpr,0,COL_W*2*dpr,FIELD_H*dpr);
  ctx.fillStyle="rgba(40,100,220,0.95)";
  ctx.beginPath(); ctx.moveTo(coolestX,(FIELD_H-2)*dpr); ctx.lineTo(coolestX-triS,(FIELD_H-2+triS*1.6)*dpr); ctx.lineTo(coolestX+triS,(FIELD_H-2+triS*1.6)*dpr); ctx.closePath(); ctx.fill();
  ctx.font="bold "+(7.5*dpr)+"px Georgia"; ctx.textAlign="center";
  ctx.fillText(COOLEST.year, coolestX, (FIELD_H-5)*dpr);
  ctx.restore();
}
// Advance one particle by one frame: apply noise force + spring tether, then draw
function stepParticle(p) {
  // Sample fBm noise field at the current particle position to get a flow angle
  const nx=p.x/W*4+t*0.005, ny=p.y/FIELD_H*4;
  const nv=fbm(nx,ny,p.seed), angle=nv*Math.PI*3;
  // Accelerate toward noise direction + spring force pulling back to home cell (xc, yc)
  p.vx+=Math.cos(angle)*p.speed*0.10+(p.xc-p.x)*0.004;
  p.vy+=Math.sin(angle)*p.speed*0.08+(p.yc-p.y)*0.018;
  p.vx*=0.90; p.vy*=0.90; p.x+=p.vx; p.y+=p.vy; p.age++; // dampen velocity, integrate position
  // Highlighted particles (selected year or hovered month) are brighter and larger
  const isHighlightYear=YEARS.indexOf(p.year)===YEARS.indexOf(highlightYear);
  const isHighlightMonth=hoverMonth!==null&&p.month0===hoverMonth;
  const isHighlighted=isHighlightYear||isHighlightMonth;
  // Fade in at birth, fade out near end of life
  const life=p.age/p.maxAge;
  const baseA=life<0.1?life*10:life>0.85?(1-life)*6.7:1;
  const alpha=Math.min(isHighlighted?0.92:0.55, baseA*(isHighlighted?0.92:0.55));
  const size=p.size*(isHighlighted?1.8:1)*devicePixelRatio;
  ctx.beginPath();
  ctx.arc(p.x*devicePixelRatio,p.y*devicePixelRatio,Math.max(0.4,size),0,Math.PI*2);
  ctx.fillStyle=anomColor(p.anom,p.isSummer,alpha); ctx.fill();
  // Recycle particle if it has aged out or drifted off canvas
  if(p.age>p.maxAge||p.x<-8||p.x>W+8||p.y<-8||p.y>FIELD_H+8){ Object.assign(p,makeParticle(),{age:0}); }
}
let t=0; // global frame counter — drives the slow drift of the noise field over time
// Main animation loop — called once per frame via requestAnimationFrame
function frame() {
  // Semi-transparent fill instead of clearRect creates a motion trail effect
  ctx.fillStyle="rgba(7,7,14,0.08)"; ctx.fillRect(0,0,canvas.width,canvas.height);
  // Subtle left-to-right color gradient reinforces the cool (1984) → warm (2024) narrative
  const bgG=ctx.createLinearGradient(0,0,canvas.width,0);
  bgG.addColorStop(0,"rgba(8,12,45,0.03)"); bgG.addColorStop(0.5,"rgba(7,7,14,0.01)"); bgG.addColorStop(1,"rgba(55,12,5,0.05)");
  ctx.fillStyle=bgG; ctx.fillRect(0,0,canvas.width,canvas.height);
  drawGrid(); particles.forEach(stepParticle); t++;
  requestAnimationFrame(frame);
}
ctx.fillStyle="#07070e"; ctx.fillRect(0,0,canvas.width,canvas.height);
boot(); updateYear(); drawBarChart(); frame();
</script>
</body>
</html>')

# ── 1. Save standalone HTML to output folder ───────────────────────────────
writeLines(particle_html_str, con = out_file)
message("Standalone HTML saved to: ", out_file)

# ── 2. Embed in Quarto document via iframe ─────────────────────────────────
# We encode the full HTML as a data URI so it works from file:// with no
# web-server requirement — bypasses the OJS web-worker restriction entirely.
b64 <- jsonlite::base64_enc(charToRaw(particle_html_str))
iframe_src <- paste0("data:text/html;base64,", b64)

htmltools::browsable(
  htmltools::tags$iframe(
    src            = iframe_src,
    width          = "100%",
    height         = "620px",
    style          = "border:none; display:block;",
    scrolling      = "no",
    allowfullscreen = NA
  )
)
```

---

## Key Findings

Forty-one years of monthly data reveal a clear and accelerating warming trend in Tucson.

::: {.callout-important}
## The Numbers

|  | Mean Temperature Anomaly |
|---|---|
| **1984–1993** (first decade) | `r paste0(round(early, 2), "°C")` |
| **2015–2024** (last decade) | `r paste0(ifelse(late > 0, "+", ""), round(late, 2), "°C")` |
| **Total shift** | `r paste0("+", round(late - early, 2), "°C")` |
| **Hottest year on record** | `r hottest$year` (`r paste0("+", round(hottest$mean_anomaly, 2), "°C")` above baseline) |
| **Coolest year on record** | `r coolest$year` (`r paste0(round(coolest$mean_anomaly, 2), "°C")` vs baseline) |
:::

The warming is not confined to summer. The monthly heat calendar and rose chart both show that anomalous warmth has spread into spring and autumn — months that once provided a thermal buffer for the city. Winters are shorter, milder, and less reliably cool than they were in the 1980s.

Tucson's blues are fading. The reds are here to stay.

---

## Session Info

```{r}
#| label: session-info
#| code-fold: false

sessionInfo()
```
